# Bias in the Data (Risk of Recidivism Analysis)

## Setup 

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, # tidyverse packages 
  conflicted, # an alternative conflict resolution strategy 
  ggthemes, # other themes for ggplot2 
  patchwork, # arranging ggplots
  scales, # rescaling 
  survival, # survival analysis
  broom, # for modeling
  here, # reproducibility 
  glue # pasting strings and objects 
)
# To avoid conflicts 
conflict_prefer("filter", "dplyr") 
conflict_prefer("select", "dplyr") 
# Set themes 
theme_set(ggthemes::theme_fivethirtyeight())
```

## Load data 

We select fields for severity of charge, number of priors, demographics, age, sex, COMPAS scores, and whether each person was accused of a crime within two years.

```{r message=FALSE}
setwd("~/Desktop")
two_years <- read_csv("compas_scores.csv")
glue("N of observations (rows): {nrow(two_years)}
      N of variables (columns): {ncol(two_years)}")
```

## Wrangling 

- Not all of the observations are useable for the first round of analysis.
- There are a number of reasons to remove rows because of missing data:
  - If the charge date of a defendants COMPAS scored crime was not within 30 days from when the person was arrested, we assume that because of data quality reasons, that we do not have the right offense.
- We coded the recidivist flag -- is_recid -- to be -1 if we could not find a COMPAS case at all.
- In a similar vein, everyday traffic offenses -- those with a c_charge_degree of 'O' -- that do not result in Jail time are removed (only two of them).
- We filter the underlying data from Broward county to include only rows representing people who either recidivated in two years, or had at least two years outside of a correctional facility.

### Create a function 

```{r}
wrangle_data <- function(data){
  df <- data %>% 
    
    # Select variables 
    select(age, c_charge_degree, race, age_cat, score_text, sex, priors_count, days_b_screening_arrest, decile_score, is_recid, two_year_recid, 
           c_jail_in, c_jail_out) %>% 
    # Filter rows 
    filter(days_b_screening_arrest <= 30,
           days_b_screening_arrest >= -30, 
           is_recid != -1,
           c_charge_degree != "O",
           score_text != 'N/A') %>% 
    # Mutate variables 
    mutate(length_of_stay = as.numeric(as.Date(c_jail_out) - as.Date(c_jail_in)),
           c_charge_degree = factor(c_charge_degree),
           age_cat = factor(age_cat),
           race = factor(race, levels = c("Caucasian","African-American","Hispanic","Other","Asian","Native American")),
           sex = factor(sex, levels = c("Male","Female")),
           score_text = factor(score_text, levels = c("Low", "Medium", "High")),
           score = score_text,
           # I added this new variable to test whether measuring the DV as a binary or continuous var makes a difference 
           score_num = as.numeric(score_text)) %>% 
    # Rename variables 
    rename(crime = c_charge_degree,
           gender = sex)
  
  return(df)}
```

### Apply the function to the data 

```{r}
df <- wrangle_data(two_years)
names(df)
# Check whether the function works as expected
head(df, 5)
```

## Descriptive analysis 

- Higher COMPAS scores are slightly correlated with a longer length of stay (time incarcerated).

```{r}
cor(df$length_of_stay, df$decile_score)
df %>%
  group_by(score) %>%
  count() %>%
  ggplot(aes(x = score, y = n)) +
  geom_col() +
  labs(x = "Score",
       y = "Count",
       title = "Score distribution")
```

>Judges are often presented with two sets of scores from the COMPAS system -- one that classifies people into High, Medium and Low risk, along with a corresponding decile score. There is a clear downward trend in decile scores as those scores increase for white defendants.

```{r}
df %>%
  ggplot(aes(ordered(decile_score))) + 
  geom_bar() +
  facet_wrap(~race, nrow = 2) +
  labs(x = "Decile Score",
       y = "Count",
       Title = "Defendant's Decile Score")
```

## Modeling 

After filtering out bad rows, our first question is whether there is a significant difference in the COMPAS scores between races. To do so we need to change some variables into factors, and run a logistic regression to compare low scores to high scores.

### Model building 

```{r}
model_data <- function(data){
  # Logistic regression model
  lr_model <- glm(score ~ gender + age_cat + race + priors_count + crime + two_year_recid, 
                  family = "binomial", data = data)
  # OLS, DV = score_num
  ols_model1 <- lm(score_num ~ gender + age_cat + race + priors_count + crime + two_year_recid, data = data)
  # OLS, DV = decile_score 
  ols_model2 <- lm(decile_score ~ gender + age_cat + race + priors_count + crime + two_year_recid, data = data)
  # Extract model outcomes with confidence intervals 
  lr_est <- lr_model %>% 
    tidy(conf.int = TRUE) 
  ols_est1 <- ols_model1 %>%
    tidy(conf.int = TRUE) 
  ols_est2 <- ols_model2 %>%
    tidy(conf.int = TRUE) 
  # AIC scores 
  lr_AIC <- AIC(lr_model)
  ols_AIC1 <- AIC(ols_model1)
  ols_AIC2 <- AIC(ols_model2)
  
  list(lr_est, ols_est1, ols_est2, 
       lr_AIC, ols_AIC1, ols_AIC2)
}
```

### Model comparisons 

```{r}
glue("AIC score of logistic regression: {model_data(df)[4]} 
      AIC score of OLS regression (with categorical DV):  {model_data(df)[5]}
      AIC score of OLS regression (with continuous DV): {model_data(df)[6]}")
```

### Logistic regression model 

```{r}
lr_model <- model_data(df)[1] %>% data.frame()
lr_model %>%
  filter(term != "(Intercept)") %>%
  mutate(term = gsub("race|age_cat|gender|M","", term)) %>%
  ggplot(aes(x = fct_reorder(term, estimate), y = estimate, ymax = conf.high, ymin = conf.low)) +
  geom_pointrange() +
  coord_flip() +
  labs(y = "Estimate", x = "",
       title = "Logistic regression") +
  geom_hline(yintercept = 0, linetype = "dashed")
```

Logistic regression coefficients are essentially logged odds ratios. "Odds" here can be expressed as  $\frac{p}{1-p}$, where __p__ is defined as a success and its complement __1-p__, as a failure. Coefficients of 1 indicate equal probability for the binary outcomes. Coefficients *greater* than 1 indicates strong chance for p and weak chance for 1-p, where coefficients *smaller* than 1 indicate the opposite. Exact interpretations of this concept is not so intuive, being that an odds of 2.0 corresponds to a probability of 1/3 (!). 

(Keep in mind that probabilities are bounded between [0, 1], where odds can range between 0 and infinity. A Log odds ranges from negative to positive infinity. We are going through all this because a logarithmic function is used in `lr_model` to map predictor variables into probabilities that fit the model with the binary outcomes we see above.)

In the next few code chunks we reinterpret coefficients by transorming logged odds ratios into another measure of association, __relative risk__.  Relative risk can be expressed as __RR = (odds ratio) / (1 - p0 + (p0 * odds ratio))__. Where p-0 is the baseline risk. 

>Relative risk is formally understood as a form of division. In this case we are measuring the risk of outcome (a higher recidivism score), relative to exposure (certain racial or other identity category).  

For more information on relative risks and its value in statistical communication, see [Grant](https://www.bmj.com/content/348/bmj.f7450) (2014), [Wang](https://www.jstatsoft.org/article/view/v055i05) (2013), and [Zhang and Yu](https://jamanetwork.com/journals/jama/fullarticle/188182) (1998). 

```{r}
odds_to_risk <- function(model){
    
    # Calculating p0 (baseline or control group)
    intercept <- model$estimate[model$term == "(Intercept)"]
    control <- exp(intercept) / (1 + exp(intercept)) 
    
    # Calculating relative risk 
    model <- model %>% filter(term != "(Intercept)")
    model$relative_risk <- (exp(model$estimate) / 
                        (1 - control + (control * exp(model$estimate)))) 
    
    return(model)
}
```

```{r}
odds_to_risk(lr_model) %>%
  relocate(relative_risk) %>%
  arrange(desc(relative_risk))
```

We see a variety of outcomes measured above. Eg: A relative risk of 1.45 (African American) indicates that black defendants are 45% more likely than our control (white male defendants) to receive a higher score.

The plot visualizes this and other results from the table. 

```{r}
odds_to_risk(lr_model) %>%
    mutate(term = gsub("race|age_cat|gender","", term)) %>% 
    ggplot(aes(x = fct_reorder(term, relative_risk), y = relative_risk)) +
        geom_point(size = 3) +
        coord_flip() +
        labs(y = "Likelihood", x = "",
             title = "Logistic regression") +
        scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
        geom_hline(yintercept = 1, linetype = "dashed")
```

